% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode,linktoc=all}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The Secret Trapdoor to Success},
  pdfauthor={R (Chandra) Chandrasekhar},
  pdflang={en-GB},
  colorlinks=true,
  linkcolor={DarkOliveGreen},
  filecolor={Purple},
  citecolor={DarkKhaki},
  urlcolor={Maroon},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=25mm]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{british}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% $HOME/.pandoc/defaults/latex-header-includes.tex
% Common header includes for both lualatex and xelatex engines.
%
% Preliminaries
%
% \PassOptionsToPackage{rgb,dvipsnames,svgnames}{xcolor}
% \PassOptionsToPackage{main=british}{babel}
\AtBeginEnvironment{quote}{\small}
\AtBeginEnvironment{quotation}{\small}
\AtBeginEnvironment{longtable}{\centering}
%
% Packages that are useful to include
%
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[inkscapeversion=1]{svg}
\usepackage[defaultlines=4,all]{nowidow}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{etoolbox}
\usepackage{fontsize}
\usepackage{newunicodechar}
\usepackage{pdflscape}
\usepackage{fnpct}
\usepackage{parskip}
  \setlength{\parindent}{0pt}
\usepackage[style=american]{csquotes}
% \usepackage{setspace} Use the <fontname-plus.tex> files for setspace
%
% noto-plus.tex
% Font-setting header file for use with Pandoc Markdown
% to generate PDF via LuaLaTeX.
% The main font is Noto Serif.
% Other main fonts are also available in appropriately named file.
\usepackage{fontspec}
\usepackage{setspace}
\setstretch{1.3}
%
\defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase,Renderer=Node} % at the start always
%
% For English
% See also https://tex.stackexchange.com/questions/574047/lualatex-amsthm-polyglossia-charissil-error
% We use Node as Renderer for the Latin Font and Greek Font and HarfBuzz as renderer ofr Indic fonts.
%
\babelfont{rm}[Script=Latin,Scale=1]{NotoSerif}% Config is at $HOME/texmf/tex/latex/NotoSerif.fontspec
%
\babelfont{sf}[Script=Latin]{SourceSansPro}% Config is at $HOME/texmf/tex/latex/SourceSansPro.fontspec
%
\babelfont{tt}[Script=Latin]{FiraMono}% Config is at $HOME/texmf/tex/latex/FiraMono.fontspec
%
% Sanskrit, Tamil, and Greek fonts
%
\babelprovide[import, onchar=ids fonts]{sanskrit}
\babelprovide[import, onchar=ids fonts]{tamil}
\babelprovide[import, onchar=ids fonts]{greek}
%
\babelfont[sanskrit]{rm}[Scale=1.1,Renderer=HarfBuzz,Script=Devanagari]{NotoSerifDevanagari}
\babelfont[sanskrit]{sf}[Scale=1.1,Renderer=HarfBuzz,Script=Devanagari]{NotoSansDevanagari}
\babelfont[tamil]{rm}[Renderer=HarfBuzz,Script=Tamil]{NotoSerifTamil}
\babelfont[tamil]{sf}[Renderer=HarfBuzz,Script=Tamil]{NotoSansTamil}
\babelfont[greek]{rm}[Script=Greek]{GentiumBookPlus}
%
% Math font
%
\usepackage{unicode-math} % seems not to hurt % fallabck
\setmathfont[bold-style=TeX]{STIX Two Math}
%
%
% Other fonts
%
\newfontfamily{\emojifont}{Symbola}
%

\usepackage{titling}
\usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead{}
    \fancyfoot{}
    \renewcommand{\headrulewidth}{0.2pt}
    \renewcommand{\footrulewidth}{0.2pt}
    \fancyhead[LO,RE]{\scshape\thetitle}
    \fancyfoot[CO,CE]{\footnotesize Copyright © 2006\textendash\the\year, R (Chandra) Chandrasekhar}
    \fancyfoot[RE,RO]{\thepage}
\newfontfamily{\regulariconfont}{Font Awesome 6 Free Regular}[Color=Grey]
\newfontfamily{\solidiconfont}{Font Awesome 6 Free Solid}[Color=Grey]
\newfontfamily{\brandsiconfont}{Font Awesome 6 Brands}[Color=Grey]
%
% Direct input of Unicode code points
%
\newcommand{\faEnvelope}{\regulariconfont\ ^^^^f0e0\normalfont}
\newcommand{\faMobile}{\solidiconfont\ ^^^^f3cd\normalfont}
\newcommand{\faLinkedin}{\brandsiconfont\ ^^^^f0e1\normalfont}
\newcommand{\faGithub}{\brandsiconfont\ ^^^^f09b\normalfont}
\newcommand{\faAtom}{\solidiconfont\ ^^^^f5d2\normalfont}
\newcommand{\faPaperPlaneRegular}{\regulariconfont\ ^^^^f1d8\normalfont}
\newcommand{\faPaperPlaneSolid}{\solidiconfont\ ^^^^f1d8\normalfont}

%
% The block below is commented out because of Tofu glyphs in HTML
%
% \newcommand{\faEnvelope}{\regulariconfont\ \normalfont}
% \newcommand{\faMobile}{\solidiconfont\ \normalfont}
% \newcommand{\faLinkedin}{\brandsiconfont\ \normalfont}
% \newcommand{\faGithub}{\brandsiconfont\ \normalfont}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{The Secret Trapdoor to Success}
\author{R (Chandra) Chandrasekhar}
\date{2023-01-29 | 2023-01-29}

\begin{document}
\maketitle

\thispagestyle{empty}


\begin{flushright}

\begin{footnotesize}

Row, row, row your boat\\
Gently down the stream.\\
Merrily, merrily, merrily, merrily,\\
Life is but a dream.\\
\href{https://en.wikipedia.org/wiki/Row,_Row,_Row_Your_Boat}{\textsc{Nursery
Rhyme}}

\end{footnotesize}

\end{flushright}

\hypertarget{from-dreams-to-real-life}{%
\subsection{From dreams to real life}\label{from-dreams-to-real-life}}

The nursery rhyme tells us that life is but a dream. Imagine, however,
if someone told you that your dreams could become your life. Would you
be inclined to believe that statement? If you are imbued with the
scientific spirit, you would at the very least subject that assertion to
experimental validation or refutation. If you are of that mind, read
this blog carefully, because it could be the single most important
\emph{secret of success}---academic or otherwise---that you might
encounter in your entire life.

\hypertarget{an-example-from-a-famous-life}{%
\subsection{An example from a famous
life}\label{an-example-from-a-famous-life}}

\href{https://en.wikipedia.org/wiki/A._P._J._Abdul_Kalam}{Dr A P J Abdul
Kalam}, eleventh President of India, in his autobiography
\href{https://en.wikipedia.org/wiki/Wings_of_Fire_(autobiography)}{\emph{Wings
of Fire}}, recounts a deeply personal episode from his life, about how
his own ambition in his early years---to be an Air Force pilot---was
shattered when he was not chosen for the role by the Selection Board.
Crestfallen, he wended his way to the ashram of the saintly monk,
\href{https://en.wikipedia.org/wiki/Sivananda_Saraswati}{Sri Swami
Sivananda} in \href{https://en.wikipedia.org/wiki/Rishikesh}{Rishikesh}.
Let us hear what transpired in his own words:

\begin{quote}
It took me some time to comprehend that the opportunity to join the Air
Force had just slipped through my fingers. I dragged myself out of the
Selection Board and stood at the edge of a cliff. There was a lake far
below. I knew that the days ahead would be difficult. There were
questions to be answered and a plan of action to be prepared. I trekked
down to Rishikesh.\\
\strut \\
I bathed in the Ganga and revelled in the purity of its water. Then, I
walked to the
\href{https://en.wikipedia.org/wiki/Divine_Life_Society}{Sivananda
Ashram} situated a little way up the hill. I could feel intense
vibrations when I entered. I saw a large number of
\href{https://en.wikipedia.org/wiki/Sadhu}{sadhus} seated all around in
a state of trance. I had read that sadhus were psychic people---people
who know things intuitively and, in my dejected mood, I sought answers
to the doubts that troubled me.\\
\strut \\
I met Swami Sivananda---a man who looked like a Buddha, wearing a
snow-white dhoti and wooden slippers. He had an olive complexion and
black, piercing eyes. I was struck by his irresistible, almost childlike
smile and gracious manner.\\
\strut \\
I introduced myself to the Swamiji. My Muslim name aroused no reaction
in him. Before I could speak any further, he inquired about the source
of my sorrow. He offered no explanation of how he knew that I was sad
and I did not ask.\\
\strut \\
I told him about my unsuccessful attempt to join the Indian Air Force
and my long-cherished desire to fly.\\
\strut \\
He smiled, washing away all my anxiety almost instantly. Then he said in
a feeble, but very deep voice, ``Desire, when it stems from the heart
and spirit, when it is pure and intense, possesses awesome
electromagnetic energy. This energy is released into the ether each
night, as the mind falls into the sleep state. Each morning it returns
to the conscious state reinforced with the cosmic currents. That which
has been imaged will surely and certainly be manifested. You can rely,
young man, upon this ageless promise as surely as you can rely upon the
eternally unbroken promise of sunrise\ldots{} and of Spring.''\\
\strut \\
When the student is ready, the teacher will appear---How true! Here was
the teacher to show the way to a student who had nearly gone astray!\\
\strut \\
``Accept your destiny and go ahead with your life. You are not destined
to become an Air Force pilot. What you are destined to become is not
revealed now but it is predetermined. Forget this failure, as it was
essential to lead you to your destined path. Search, instead, for the
true purpose of your existence. Become one with yourself, my son!
Surrender yourself to the wish of God,'' Swamiji said.
\protect\hyperlink{ref-kalam-wof-1999}{{[}1{]}},
\protect\hyperlink{ref-kalam-wof-online}{{[}2{]}}
\end{quote}

\hypertarget{books-on-manifesting}{%
\subsection{Books on ``manifesting''}\label{books-on-manifesting}}

There has been a steady procession of books on ``manifesting'' desires.
``Manifesting'' means exerting by doing something to make your desires
come true in the physical reality we inhabit.

Wings of Fire https://ati.dae.gov.in/ebooks.html
https://ati.dae.gov.in/ati12052021\_8.pdf: Wings of Fire online. Swami
Sivananda's quote. भारत सरकार/Government of India परमाणु ऊर्जा
विभाग/Department of Atomic Energy प्रशासनिक प्रशिक्षण संस्‍थान/Administrative
Training Institute
https://en.wikipedia.org/wiki/Wings\_of\_Fire\_(autobiography)

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\hypertarget{feedback}{%
\subsection{Feedback}\label{feedback}}

Please \href{mailto:feedback.swanlotus@gmail.com}{email me} your
comments and corrections.

\noindent A PDF version of this article is
\href{./feeling.pdf}{available for download here}:

\begin{normalsize}

\begin{ttfamily}

\url{https://swanlotus.netlify.app/blogs/feeling.pdf}

\end{ttfamily}

\end{normalsize}

\%\% \% SAS Chapter Boilerplate \% \% Chapter name: Feeling Is the
Secret \% \% File name: Feeling.tex \% \% Description: Based on Neville
Goddard's \emph{Feeling Is the Secret} \% \% History: \% \% First
written: 2023-01-12 \% Last revised : 2023-01-16 \% \%\%

\% \% Master key book \% \textit{Feeling Is the Secret}\} \% \% YouTube
video features an audio-video book: it isr read aloud and you can also
see the pages. \% \% https://www.youtube.com/watch?v=ffNWoefuwPM \%

\{As a student, you want to excel in your studies by putting in the
effort, and reaping success at the examinations. It is the mismatch
between expectations and results that leads to the slippery slope of
academic failure, which robs you of the exhilaration of success, and
sometimes even of the desire to achieve.

Suppose someone told you that the world that you encounter externally is
really a reflection of the world you harbour within yourself in your
mind. And, moreover that frustration with failure, and the futility of
all effort, may be flipped completely, simply by changing the internal
worldview in your mind.

Sounds too good to be true? What have you to lose? Try the method
outlined in this chapter and see for yourself. I have tried it and it
has worked for me as long as I played by the rules of the game.

I am deliberately not disclosing any more, to whet your appetite all the
more. There is a video presentation that delivers the theory and
practice behind this almost magical method. Try it and enjoy a smoother
ride to academic success.\}

\hypertarget{bibliography}{%
\section*{References}\label{bibliography}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-kalam-wof-1999}{}}%
\CSLLeftMargin{{[}1{]} }%
\CSLRightInline{APJ Abdul Kalam and A Tiwari, \emph{{Wings of Fire: An
Autobiography}}. Universities Press, 1999. }

\leavevmode\vadjust pre{\hypertarget{ref-kalam-wof-online}{}}%
\CSLLeftMargin{{[}2{]} }%
\CSLRightInline{APJ Abdul Kalam and A Tiwari, {`{Wings of Fire: An
Autobiography}'}, 2016. {[}Online{]}. Available:
\url{https://ati.dae.gov.in/ati12052021_8.pdf}. {[}Accessed:
29-Jan-2023{]}}

\end{CSLReferences}



\end{document}
