% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode,linktoc=all}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=25mm]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{british}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% $HOME/.pandoc/defaults/latex-header-includes.tex
% Common header includes for both lualatex and xelatex engines.
%
% Preliminaries
%
% \PassOptionsToPackage{rgb,dvipsnames,svgnames}{xcolor}
% \PassOptionsToPackage{main=british}{babel}
\PassOptionsToPackage{english}{selnolig}
\AtBeginEnvironment{quote}{\small}
\AtBeginEnvironment{quotation}{\small}
\AtBeginEnvironment{longtable}{\centering}
%
% Packages that are useful to include
%
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[inkscapeversion=1]{svg}
\usepackage[defaultlines=4,all]{nowidow}
\usepackage{etoolbox}
\usepackage{fontsize}
\usepackage{newunicodechar}
\usepackage{pdflscape}
\usepackage{fnpct}
\usepackage{parskip}
  \setlength{\parindent}{0pt}
\usepackage[style=american]{csquotes}
% \usepackage{setspace} Use the <fontname-plus.tex> files for setspace
%
\usepackage{esdiff} % for derivative symbols
\usepackage{amsmath}
\usepackage{hyperref} % cleveref must come AFTER hyperref
\usepackage[capitalize,noabbrev]{cleveref} % Must come after hyperref
% noto-plus.tex
% Font-setting header file for use with Pandoc Markdown
% to generate PDF via LuaLaTeX.
% The main font is Noto Serif.
% Other main fonts are also available in appropriately named file.
\usepackage{fontspec}
\usepackage{setspace}
\setstretch{1.3}
%
\defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase,Renderer=Node} % at the start always
%
% For English
% See also https://tex.stackexchange.com/questions/574047/lualatex-amsthm-polyglossia-charissil-error
% We use Node as Renderer for the Latin Font and Greek Font and HarfBuzz as renderer ofr Indic fonts.
%
\babelfont{rm}[Script=Latin,Scale=1]{NotoSerif}% Config is at $HOME/texmf/tex/latex/NotoSerif.fontspec
%
\babelfont{sf}[Script=Latin]{SourceSansPro}% Config is at $HOME/texmf/tex/latex/SourceSansPro.fontspec
%
\babelfont{tt}[Script=Latin]{FiraMono}% Config is at $HOME/texmf/tex/latex/FiraMono.fontspec
%
% Sanskrit, Tamil, and Greek fonts
%
\babelprovide[import, onchar=ids fonts]{sanskrit}
\babelprovide[import, onchar=ids fonts]{tamil}
\babelprovide[import, onchar=ids fonts]{greek}
%
\babelfont[sanskrit]{rm}[Scale=1.1,Renderer=HarfBuzz,Script=Devanagari]{NotoSerifDevanagari}
\babelfont[sanskrit]{sf}[Scale=1.1,Renderer=HarfBuzz,Script=Devanagari]{NotoSansDevanagari}
\babelfont[tamil]{rm}[Renderer=HarfBuzz,Script=Tamil]{NotoSerifTamil}
\babelfont[tamil]{sf}[Renderer=HarfBuzz,Script=Tamil]{NotoSansTamil}
\babelfont[greek]{rm}[Script=Greek]{GentiumBookPlus}
%
% Math font
%
\usepackage{unicode-math} % seems not to hurt % fallabck
\setmathfont[bold-style=TeX]{STIX Two Math}
%
%
% Other fonts
%
\newfontfamily{\emojifont}{Symbola}
%

\usepackage{titling}
\usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead{}
    \fancyfoot{}
    \renewcommand{\headrulewidth}{0.2pt}
    \renewcommand{\footrulewidth}{0.2pt}
    \fancyhead[LO,RE]{\scshape\thetitle}
    \fancyfoot[CO,CE]{\footnotesize Copyright © 2006\textendash\the\year, R (Chandra) Chandrasekhar}
    \fancyfoot[RE,RO]{\thepage}
\newfontfamily{\regulariconfont}{Font Awesome 6 Free Regular}[Color=Grey]
\newfontfamily{\solidiconfont}{Font Awesome 6 Free Solid}[Color=Grey]
\newfontfamily{\brandsiconfont}{Font Awesome 6 Brands}[Color=Grey]
%
% Direct input of Unicode code points
%
\newcommand{\faEnvelope}{\regulariconfont\ ^^^^f0e0\normalfont}
\newcommand{\faMobile}{\solidiconfont\ ^^^^f3cd\normalfont}
\newcommand{\faLinkedin}{\brandsiconfont\ ^^^^f0e1\normalfont}
\newcommand{\faGithub}{\brandsiconfont\ ^^^^f09b\normalfont}
\newcommand{\faAtom}{\solidiconfont\ ^^^^f5d2\normalfont}
\newcommand{\faPaperPlaneRegular}{\regulariconfont\ ^^^^f1d8\normalfont}
\newcommand{\faPaperPlaneSolid}{\solidiconfont\ ^^^^f1d8\normalfont}

%
% The block below is commented out because of Tofu glyphs in HTML
%
% \newcommand{\faEnvelope}{\regulariconfont\ \normalfont}
% \newcommand{\faMobile}{\solidiconfont\ \normalfont}
% \newcommand{\faLinkedin}{\brandsiconfont\ \normalfont}
% \newcommand{\faGithub}{\brandsiconfont\ \normalfont}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{sf}
\hypersetup{
  pdftitle={The Most Scary Experience},
  pdfauthor={R (Chandra) Chandrasekhar},
  pdflang={en-GB},
  colorlinks=true,
  linkcolor={DarkOliveGreen},
  filecolor={Purple},
  citecolor={DarkKhaki},
  urlcolor={Maroon},
  pdfcreator={LaTeX via pandoc}}

\title{The Most Scary Experience}
\author{R (Chandra) Chandrasekhar}
\date{2023-07-14 | 2023-07-24}

\begin{document}
\maketitle

\thispagestyle{empty}


\hypertarget{probing-fear}{%
\subsection{Probing fear}\label{probing-fear}}

My redoubtable friend Solus ``Sol'' Simkin wandered into my office late
one afternoon and asked me, ``What is the most scary experience for a
human being?''

I thought for an instant as I replied, almost reflexively, ``Death. What
else? Or a close shave with death.''

The hard taskmaster that he is, Sol told me to think again. He told me
to imagine something that pursues one's days and nights like a
relentless shadow. A fear with no respite. Not the fear of encountering
a cobra that wears out in half an hour at most. Not the fear of a
narrowly missed road accident that leaves one rattled and shivering for
a full five minutes. No, this is a shoreless fear. And nothing
supernatural. It is a fear as mundane as the Earth and yet, it is a fear
that we do not normally encounter, let alone experience. But when
recognized and felt, it is a fear that chills the spine and shakes one's
very core.

``Sol,'' I said. What you are asking me to probe calls for long, deep,
and hard thought. I need more time''.

\hypertarget{the-genre-of-horror-fiction}{%
\subsection{The genre of horror
fiction}\label{the-genre-of-horror-fiction}}

I had hastily tried to assemble a quick list of scary tales from
\href{https://americanliterature.com/author/hh-munro-saki}{H H Munro's}
\href{https://www.classicshorts.com/stories/vashtar.html}{\emph{Shredni
Vashtar}}, to \href{https://www.hplovecraft.com/}{H P Lovecraft's}
\href{https://www.hplovecraft.com/writings/texts/fiction/cc.aspx}{\emph{The
Call of Cthulhu}}. Then there was
\href{https://openlibrary.org/books/OL4122966M/The_Arbor_House_treasury_of_horror_and_the_supernatural}{\emph{The
Arbor House Treasury of Horror and the Supernatural}} that engrossed me
for many days and nights with its eclectic collection of terrifying
tales, but none of them had left me a helpless heap of jelly, paralyzed
with fear. Then I remembered my one of my favourite authors,
\href{https://www.poetryfoundation.org/poets/edgar-allan-poe}{Edgar
Allan Poe}, and his
\href{https://www.amazon.in/Tales-Mystery-Imagination-Collins-Classics/dp/0007420226}{\emph{Tales
of Mystery and Imagination}}, surely among the choicest morsels of the
macabre. The injection of fear from these stories had been swift and
intense, but not long-lasting. Even the movie,
\href{https://www.imdb.com/title/tt0081505/}{\emph{The Shining}}, based
on angst-meister
\href{https://www.britannica.com/biography/Stephen-King}{Stephen King's}
novel of the same name, had been able to terrify me but fleetingly.

``Take your time, and wend your way slowly through the labyrinth of your
own memories and pluck for me that single, shiny nugget of fear that
dazzles you even today, its power to enchant , enthrall and engulf,
undiminished by the passage of time,'' Sol told me soothingly. ``I have
a reason for asking you to undertake this introspection. I had my own
close brush with depthless fear recently. What I learned from it was
unusual to say the least. I wish to discuss it with you. So, when you
are ready, we will discuss it in the relaxing ambiance of our favourite
coffee hangout. Mind you, be prepared for many sessions before the
denouement takes place.''

\url{https://www.302aw.afrc.af.mil/News/Commentaries/Display/Article/191174/the-prison-of-hopelessness/}

Help O Power Eternal
\url{https://www.gutenberg.org/files/64988/64988-h/64988-h.htm}

\begin{quote}
``Why is the word `hope' magic?'' asked the youngest camel, stretching
out one stiff leg to see if it still could move. And now Mohammed's son
lifted the little camel's head up again and laid it against his shoulder
while he shook the remaining cords away. When he did this, the little
camel saw that he was young and very handsome. He was wearing a silk
turban with pearls and turquoises embroidered on it, and carved gold
ornaments{[}35{]} hung from his ears, and there was a look of great
gentleness in his face.
\end{quote}

\begin{quote}
``Well, you see, h stands for `help,' and o stands for `O,' and p stands
for `power,' and e stands for `eternal,'\,'' he said so lightly and
merrily that he seemed to be making fun of something. He took out a
little ivory flask from his garments and poured some fresh water between
the little camel's burning lips. ``So when you say `hope' like that,
you're really saying `Help, O power eternal!' And that means me because
I've been appointed your patron saint this year.''
\end{quote}

Unfinished business

PhD-release syndrome

Post-partum depression

Sri Ramaksishna

Unfinished symphonies/unfinished business

The scariness of the Void

Networking: We are all connected

We are Nature's Children: smell the roses; smell the coffee; listen to
birdsong etc.

Be part of something greater than yourself: serve others

Ultimately, live life with gusto so that joie de vivre spreads from you
to others like a sunshine laden zephyr

Poem on ``Purpose'' from my book

Venkata Vara Prasad episode

Loneliness versus Solitude

Only swimmer in the swimming pol; the last thought; the immensity of the
void

\url{https://www.youtube.com/watch?v=Iy99Zhp24LY}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\hypertarget{feedback}{%
\subsection{Feedback}\label{feedback}}

Please \href{mailto:feedback.swanlotus@gmail.com}{email me} your
comments and corrections.

\noindent A PDF version of this article is
\href{./most-scary-experience.pdf}{available for download here}:

\begin{small}

\begin{sffamily}

\url{https://swanlotus.netlify.app/blogs/most-scary-experience.pdf}

\end{sffamily}

\end{small}



\end{document}
